\documentclass[12pt, a4paper]{article}
\usepackage[utf8]{inputenc}

\usepackage{graphicx}

\usepackage{geometry}

\usepackage{multicol}
\usepackage{listings}


\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}

\geometry{margin=0.6in}


\setlength{\parindent}{0em}
\setlength{\parskip}{1em}

\title{NetSec}

\begin{document}
\section{Introduction}
the terms information, computer and network security get confused, but the difference is important
\begin{itemize}
    \item information security regards all the the prevention of any damage involving data, in any format
    \item Computer security regards the protection of the phisical machine and the data it contains
    \item Network security regards the protection of the network and infrastructure of a company
\end{itemize}
cybersecurity is the ability to protect the use of cyberspace (internet, private networks,...) from cyber
attacks

\subsection{Remarks on security}
\subsubsection*{Adversaries}
Security is characterized by protection against malicious adversaries, with different motivations (money
or glory), different capabilities and objectives. This details compone the modelling of a malicious 
adversary

\subsubsection*{Mitigation}
To mitigate a threat, various strategies are aviable, regarding people, processes and technologies.
This controls are classified in preventive, detective and corrective.\\What conditionates which tactic 
to select is risk management, or the process to identify the probability of an attack and the extention 
of the subsequent damage

\subsubsection*{Trust}
A library, or dependency, or every third party product to a software, requires trust and dependability,
also known as the ability to avoid failures.

\subsubsection*{Residual Risks}
All security controls can mitigate an attack, but do not avoid it completely. In some cases they can contain more vulnerabilities

\subsection{Cia Triad}
The most important security properties are Confifentiality, Integrity and Availability

\subsubsection*{Confidentiality}
The main objective of confidentiality is to protect personal privacy and proprietary information from unauthorized indiviaduals.
Confidentiality covers data in storage, during processing and while in transit

\subsubsection*{Integrity}
In this case our objective is to avoid unauthorized modification or destruction of data, including the authenticity and non-repudiation

\subsubsection*{Availability}
Ensuring timely and reliable access to data and services by authorized users

\subsection{Risk}
\subsubsection*{Vulnerability}
A vulnerability is a weakness or flaw in a system, procedure, design or network that can be exploited by a threat source.
They are characterized by the difficulty in identifying and exploiting them

\subsubsection*{Threat}
A threat is any circumstance or event that could impact organiazational operations or assets via unauthorized access, destruction and 
modification of data or denial of service. They are characterized by the propensity to attack (intent) and the ability to successfully do it (capability)

\subsubsection*{Risk}
the probability that a threat will exploit a system vulnerability is called risk, and it's calculated in function of the impact and 
the likelihood of occurrence.

The risk is anilized though a matrix $5x5$ which uses the the severity of the impact and the likelihood to happen as axis and gives point based on the
multiplication of the axes

\subsection{Secutity policy, service and mechanism}
This three things are, in order, the rules set by an organization to protect data, the capabilities to support one or more security requirements and the
devices and functions used to protect data and provide security services

\newpage
\section{Security in perspective}
\subsection{Attacks and impact}
Every component of the CIA triad is an objective of attacks. The violation of confidentiality leads to a disclosure of your personal
information. In the integrity case impacts on the trustworthiness of data and the violation of availability brings economic damage

Further difficulties come from the ever extending attack surface and the acquisition of cyber-intelligence information. The second case 
is an issue because corporations don't usually share informations about their security system or the vulnerabilities they found.

\subsection{Security and privacy}
They are two different but overlapping properties. Privacy can be also used to indicate confidentiality or controlled sharing.

\subsubsection*{Non-repudiation}
It's a protection against an individual falsely denying to having performed a particular action. The opposite is plausible 
deniability. This two cases are a conflict of privacy and security because one violates the other.

\subsection{Legal constraints}
The GDPR (General Data Protection Regulation) is a regulation that applies all european companies and non european companies
that process europeans personal data. It protects all the data regarding genetic, metal, cultural, economic and social identities.
When it's not followed it applies penalities like fines up to 20 million euros or the 4\% of annual global revenue of a company.
The GDPR requires that products, systems and processes follow privacy-by-design concepts and a risk-based approach to cybersecurity 

\subsection{Security and human factors}
The human factor can be a great vulnerability in a system, so it's important to provide security training for the employees.
The usual concept to follow when creating security policies for the human factor is:
\begin{center}
    Make it easy to do the right thing, hard to do the wrong thing and easy to recover when the wrong thing happens
\end{center}

\newpage
\section{Authentication - Passwords and more}
Passwords are the main human computer authentication system, but more secure and userfriendly alternatives are researched.
In a secure system  the user identity is used as a parameter for access controll and all relevant events are logged.

The current situation has found that there isn't a perfect authentication system, but a combination of password, OTP and other
techniques is more likely to succeed

\subsection{Lessons learned}
Passwords were created for protection against jokes and abuse of resources, but it was easy to guess passwords or to find the 
passwords stored in clear. To prevent this password hashing and salting were invented. After a major attack (morris worm) password
were stored in a shadowfile readable only by the root user

\subsection{Passwords and the web}
With the arrival of the internet, new issues with passwords have been discovered: the reuse of the same password for different 
services, phishing and the need to reset forgotten passwords. They have been made some attempts to outsource the authentication
to third party services, and smartphones increase security using the two factors authentication.

\subsection{User authentication and digital identity}
When logging on to a computer you enter username (announcing who you are) and password (proving to be who you claim to be). This
type of authentication is called \textbf{user authentication}, in this case, \textbf{by password}.

BOOTSTRAPING

The process of identity proofing follows 3 steps:
\begin{itemize}
    \item Resolution: collection of data/evidence called Personal Identifiable Information (PII)
    \item Validation: processing of the evidence by an authoritative source, searching for coherence of the information and 
    complaiance against standards
    \item Verification: the evidence is verified and the user is allowed to proceed
\end{itemize}
There are different levels of certainty of one user identification, these are called Identity Assurance Levels (IALs), and 
are divided in 3 levels. With IAL 1, all attributes are self-asserted by the user or should be treated as self-asserted.
at level 2 requires either remothe or in-person identification. Lastly, at third level, is required the verification by
an authorized representative.

\subsection{Attacks and mitigations}
One way to get a password is to guess it. This opens the way to two attack types: the brute force, which tries all possible 
combinations of symbols, and the smarter dictionary, which uses a list of commonly used passwords.\\This type of attacks
can't be prevented, but the probability to be successful can be reduced. The mitigations are:
\begin{itemize}
    \item change default passwords
    \item forbid commonly used passwords
    \item limit the number of attempts
    \item don't use password hints or knowledge based
    \item incourage the use of password generators and managers
\end{itemize}

\subsubsection*{Spoofing}
Password spoofing is the use of programs with fake login interfaces that captures user and password when a legittimate user 
gets bamboozled. The countermeadure is to implement mutual authentication between user and OS

\subsubsection*{Phishing}
an attacker impersonates the system to trick a unser into releasing sensible data. It can also use social engineering 

\subsubsection*{Credential stuffing}
Is the use of information from compromised databases to be applied on other services

\subsubsection*{Password file protection}
To maintain confidentiality and integrity the file could use cryptographic protection, access controll enforced by the OS,
and further measures to slow dictionary attacks

\subsection{Hashing and salting}
The Hash is a one way function easy to compute but hard to reverse. Password usually get stored in the form of hash.
The characteristics of an hash function are the easy computation, compression, one-way and collision resistance.

NB: strong collision resistance is the property to be improbable to produce 2 messages with the same hash code. This can 
be problematic if your storing passwords. Weak collision is when two passwords retrieve the same hash in the login. This 
could lead to an exploit of the system security

The principal algoritms in use are RIPEMD-160 and SHA-256 (others have been cracked like MD4, MD5, SHA-1). 
Hashing can be broken through the use of dictionary or rainbow (hash dictionaries) attacks.

To increase the security salting has been introduced to the hashes. Salting is the practice of appending random values
to the end of the password before hashing

\subsection{Extensions for password based authentication}
Multi factor authentication is characterized by the addition of some properties like knowledge, possession or inherence.
The most common is two factor authentication (2FA) 

\subsubsection{Time based one time passwords}
it uses an algoritm to generate the same number on the server and the host, then the user sends the value that is shown 
by the host, and if the two number are equals, the user is authorized. This methods still hase some weaknesses, like the
value could be phished and replayed, or the algoritm could be stolen and used by the attacker to create new OTP

Smartphones are widely used in MFA, they can provide additional mobile systems to generate password codes, but they also 
represent vulnerabilities thanks to weak protection of text messages, SIM cloning and the event that the smartphone gets 
stolen

\subsection{Assurance level (NIST)}
\begin{enumerate}
    \item requires at least single factor authentication, provides some assurance the claimant controls the authenticator
    \item requires two factor authentication, provides high confidence that the claimant controls the authenticator
    \item requires proof of possession, providing very high confidence that the claimant controls the authenticator
\end{enumerate}

\subsection{Outsourcing authentication}
Securing efficiently the authentication on a system cn be difficult for smaller organizations, so he solution is to 
delegate authentication to trusted third party providers. 

Single Sign On (SSO) is one method the uses outsourcing. It's a mechanism that collects your password on the first
login, and automatically repeats it on the successives logins. This increases the usabilty of a system and allows the 
use of more complex password, at the cost of just one password to compromise.\\This mechanism is also an example of the balance
between convenience and security

\newpage
\section{Cryptography}
\subsection{Definitions}
\begin{itemize}
    \item Cryptography is the science and study of secret writing
    \item Cryptanalysis is the science and study of methods of breaking ciphers
    \item Cryptology is the combination of cryptography and cryptanalysis
    \item Cryptosystem is the combination of:
    \subitem $E$ is an encryption algoritm
    \subitem $D$ is a decryption algoritm
    \subitem $M$ is a set of plain texts
    \subitem $K$ is the set of keys
    \subitem $C$ is the set of ciphertexts
    \subsubitem we can define $E$ and $D$ as $E:M\times K\to C$ and $D:C\times K\to M$
    \item Kerckhoff's principle: the algoritm should not be the secret, but the key
\end{itemize}

\subsubsection{Keys}
Are one input to a cryptographic algorithm and the most important part to keep secret. the set of all the possible keys
is called keyspace, and the variance of a key is measured with entropy. Key are important because they're computationally
secure 

During a connection is common to change the key to limit the amount of data that could be compromised with a single key.
So having a strong distribution system is essential 

Cryptography is just one part to make a system secure, the keys must stored securely

\subsection{Encryption}
The encryption process can be applied on blocks of same lenght or stream of data. The transformation could be through:
\begin{itemize}
    \item Substitution: each element is mapped into another element
    \item Transposition: elements are rearranged
\end{itemize}

\subsubsection{Caesar cipher (ROTk)}
It's an algoritm based on rotation, matematicaly:
\begin{center}
    $E(x)=(x+k)(mod\ 26)\qquad x\in P$\\
    $D(x)=(x-k)(mod\ 26)\qquad x\in C$
\end{center}
it's one of the easiest cipher to break

\subsubsection{Vigenere cipher}
it's an evolution of the Caesare cipher, it uses the same principle but instead of rotating with the same value, it's
used a repeated key

\subsubsection{Columnar cipher}
The message is written on rows of a table with fixed column lenght defined by the key (the empty spaces are filled with
nulls). In the end the columns are scrambled based on the key and the message gets rewritten from each column

\subsection{Modern encryption}
With the invention of computers, the algoritms had to evolve because they were too easy to brute force. In this way, new 
methods arise 

\subsection{Symmetric keys}
A single key is used to encrypt and decrypt and is share between all intended receivers. it is categorized in two types: 
stream cyphers and block cyphers. One detail on block cyphers they use both substitution an transposition through S-boxes
to obtain confusion and P-boxes to obtain diffusion. This two properties are responsable of the dependency of the key and 
the fact the if we change one bit, the entire message cyphertext changes

\subsubsection*{Stream ciphers}
converts plaintext one bit at a time with xor and the key. it's more efficient and it's used for realtime applications

\subsubsection*{Block ciphers}
breaks the message in successive blocks and enciphers them with the same key

\subsubsection{Feistel}
also called DES, it's a block cipher and follows these procedure
\begin{enumerate}
    \item splits the blocks in two
    \item applies substitution to one half, and the output is xored with the other half
    \item it swaps the two halves
\end{enumerate}

\subsubsection{AES}
it's the successor to DES. The encryption process is based on a series of table lookups and xor operations. The difference 
with DES is that the encryption and decryption algorithms are separately implemented, even if closely related. It also
mounts larger keys

\subsection{Asymmetric keys}
It's the base for a secure communication without sharing a secret key. It uses one way functions, those are function
easy to compute but difficult to reverse. This is also supported by trapdoor oneway functions, which use a key to 
easily invert the original function.

During the communication are used two keys, one to encode and another to decode. The two keys are different but 
mathematically related, although one key doesn't allow to determine the other one.\\The public key is available
to everyone the owner wants, and the private key is kept secret. Then the sender encrypts the message with the pubblic 
key of the receiver, which decrypts using his private key.

This mechanism can also be used as a digital signature. A private key is unique, so a host can encode something with his
private key so that the receiver can verify using the sender public key.This also ensures non repudiation 
that the receiver 

\subsubsection{RSA}
Is the most used in key exchange, digital signature and encryption of small blocks of data. It uses variable sized 
encryption blocks and a variable size key. As a mathematical trick it uses the factorization to prime numbers

See FMI

\subsubsection{DH}
It uses 
BALZO TATTICOù

\newpage
\section{PKI and TLS}
The public key crypto are more secure and supply high levels of integrity, confidentiality and non repudiation.
All at the cost of efficiency and the lacking of authentication. PKC is the solution to this cons.

\subsection{Digital signature}
Some data that vouches the origin and integrity of a message, it uses the private key to "sign" the message so that 
the recipient can use the public key of the sender to verify the origin. This system used hash fucntion to enable 
the signature, but today hash functions are considered deprecated, plus this method still doesn't guarantee the 
real identity of the user

\subsection{Public Key infrastructure}
The main requirements on the public key are to be bound on the identity of the party controlling the private key and
that this bound is still valid. These two requirements can be satisfied via PKI.

The initial proposal to bound public keys and identities were bulletin boards. In IoT the public key were hardcoded
in the software and in internet applications is widely spread the use of digital certificates signed by a Trusted 
Third Party (TTP)

The PKI binds the entity to the public key through a process of registration and issuance by a Certificate 
Authority (CA). Then is the Registration Authority that assures a valid and correct registration. The last component
of the PKI is the Validation Authority who provides informations of the entity on behalf of the CA

\subsubsection{Digital Certificates}
This certificates bind a public key and a person/service/hardware component/... and it contains the issuer, the 
subject (or user), the public key and the digital signature of the issuer.

The certificates are stored in a certificate distribution system, togheter with the Certificate Revocation List
(CRL)

Proccedure to obtain a certificate:
\begin{enumerate}
    \item The user generates a public and private key-pair or is assigned one by some authority
    \item The user requests the certificate to the CA server
    \item The CA answers with the certificate including public key and digital signature
    \item The user gathers the information required by the CA
    \item The user send a certificate request (CSR) to the CA including public key and additional data. The
    request is digitally signed by the user 
    \item The CA verifies the identity of the user and generates a certificate binding the user identity with 
    his public key
    \item The CA issues the certificate signed by the CA to the user
\end{enumerate}

\subsubsection{Requirements on PKI}
\begin{itemize}
    \item The TTP must be able to check a party identity
    \item The relying parties shall be able to check the time and general validity of the certificate
    \subitem One way to do this is through CRLs, the other is through the OCSP, or 
    Onlne Certificate Status Protocol
    \item The cryptographic software of the relying party must be updated according to the latest known
    vulnerabilities. The software validating certificates shall also work correctly
\end{itemize}

\subsection{SSL and TLS}
The Secure Sockets Layer and his successor the Transport Layer Security are protocols developed to secure
comunications between client and server

\subsubsection{TLS in the browser}
His goal is to provide the user with identity of page origin. HTTPS (HTTP + TLS) provides the authentication
of the web server and bidiractional encryption against man-in-the-middle, eavesdropping and tampering attacks
while protecting privacy and integrity of exchanged data

\subsubsection{TLS overview}
TLS is composed of two main protocols: the handshake protocoll which consist in the use of the public key
criptography tto share a secret simmetric key between client and server, and the record protocol which uses
the key obtained through the handshake to encrypt the communication. It also uses some additional protocols 
like the Change Cipher Protocol to switch to simmetric key encryption, and the Alert Protocol to report 
failures

The handshake protocol step by step:
\begin{enumerate}
    \item The clients sends a hello message containing the supported cipher suite
    \subitem A cypher suite is the set of algoritms used to secure a network connection
    \subsubitem They usually include key exchange, bulk encryption and message authentication code algoritms
    \subsubitem it also can include signatures and authentication algorithms to authenticate server and clint 
    \item The server answers with a hello message containg the chosen protocol and cipher suite, plus the session ID 
    \item IF the client has requested an authenticated connection, the server must send an X.509 certificate.
    \subitem it is also the step with the server key exchange
    \item the client verifies the server certificate and answers with his own certificate and his part in the
    client key exchange
    \subitem the key exchange contains the pre-master key which is used to calculate the master key
    \subitem the master key is obtainad using a pseudo random function 
    \item The server sends a change cipher spec, which determins the beginning of the use of the new cipher
    \subitem it's also sent the finished, or an hash generated from the entire handshake, used to signal the
    completition of the algoritm
    \item At this point every message is sent using the shared key
\end{enumerate}

TLS provides authentication for both server and client by encrypting the clint's choosen key with the server public
key, then the server uses the public key of the client to decrypt the data sent by the client in phase 4. If this 
exchange fails, the session terminates.\\This entire communication is encrypted, this provides confidentiality 
within the session.\\Lastly TLS uses a Message Authentication Code (MAC) to provide data integrity. The MAC uses
an algorithm composed of cryptographic functions which are similar to hash functions and digital signatures but
use differetn security requirements

\subsubsection{TLS Vulnerabilities}
TLS suffers in securities issues because of logical flaws, compatibility with old cipher suites and implementation
issues

RC4NOMORE: An attack born in 2017, it used guesses and observation of occurences of the cipher RC4 to authenticate 
as the victim

Poodle: an exploit of SSL 3.0. The attacker impersonates the server to downgrade the connection to SSL 3.0, then 
uses a vulnerability of the protocol to break the entire thing

Bleichenbacher attack: a vulnerability of SSL, it guesses the padding by looking at the server response. In TLS
is fixed as decryption failures are hidden from the attacker

Heartbleed: find in the extension heartbeat which kept connections open. The message was composed of the data and 
datalenght, from which the server replayed with the same message. But if the data was shorter than the declared 
lenght, the server would fill the missing space with data from his memory, which could have contained sensible 
information

All this vulnerabilities were using the weakness of the ciphers or failures in the logic, which in successive 
updates were patched. Right now the best mitigations for TLS vulnerabilities are proper configuration of the 
TLS server

\subsubsection{TLS 1.3}
This new version containes:
\begin{itemize}
    \item Clean-up: removal of unsafe or unused feature, like a smaller cipher suites Poodle
    \item Security: application of modern techniques
    \item Privacy: more extensive encryption
    \item Performance: Faster handshake (1 RTT and 0 RTT)
    \item Backwards compatability
\end{itemize}

NB: forward secrecy or perfect forward secrecy is the property that session keys will not be compromised even
if long-term secrets are compromised

\subsubsection{Ephemeral Diffie-Hellman}
This variant of the normal DH from the fact that DH uses the same private key. In EDH, a temporary DH key is 
generated for every connection. On the other side this doesn't allow for authentication, in fact, in TLS is used 
Authenticated Encryption with Associated Data (AEAD) cyphers.

\newpage
\section{Single Sign On, SAML, SPID, CIE}
\subsection{SSO}

SSO operates inside of security domains, or applications trusting a common security token for authentication.\\
The basic concept is:
\begin{enumerate}
    \item an user access the application
    \item the service provider refers for authentication with the identity provider
    \item the identity provider ask the user for credential and after the user is authenticated, provides 
    authentication evidence to the service provider
    \item the service provider gives to the user an authentication token
\end{enumerate}

Using this system, the credential never leave the authentication domain, but the service provider must trust 
the authentication domain, plus the authentication transfer must be protected

\subsection{SAML}
The Security Assertion Markup Language is a common language and flow between systems that want to provide an SSO 
experience, and it's XML based.

SAML distinguishes two main entities: and identity provider, who authenticates the user and provides authorisation 
information, and the service provider, which relies on the information provided by the IsP. IdP e SP share 
metadata in any form and mean possible, sharing at least the entity ID and the cryptographic keys.\\
There are also federations, or groups of entities, which establish the initial trust among resources.

The authentication flow follows this procedure:
\begin{enumerate}
    \item A user tries to access an SP
    \item The user is redirected to a Discovery Service
    \subitem it allowes the user to choose the IdP
    \item The user goes back to the SP with the ID of the IdP
    \item The user is redirected to the IdP
    \item the Authentication is performed
    \item THe user goes back to the SP with the authentication
\end{enumerate}

NB: SAML can support other resources, and resources can support multiple SAML profiles. The most used is 
redirect

SAML is composed of the authentication context, metadata and the main content, which is it self composed 
of profiles, bindings, protocols and assertions.

\subsubsection*{Assertions}
The assertion are a set of statements made by the authority,
and is classified in authentication assertion (which describes the issuer, the issued, validity period,...), 
attribute assertion (defining specific details about the subject) and authorization assertion (defines what the
subject is entitled to do)

\subsubsection*{Protocols}
This section defines the communication and cryptographic protocols used for the exchange

\subsubsection*{Bindings}
SAML uses a mechanism called SAML binding to transport the messages between the actors. There are various types 
like SAML URI, HTTP redirect, HTTP POST,...

\subsubsection*{Profiles}
The combination of protocols, assertions and binding define a profile which creates a federation and enables 
federated SSO. This profiles can be Web browser SSO, Single Logout, artifact resolution...

The two types of web SSO are \\ IsP initiated SSO
\begin{enumerate}
    \item A user is challenged to supply credentials to the IdP site
    \item The user provides credentials and a local security context is created
    \item The user accesses to the SP through the IdP, which triggers the SSO service to be called
    \item The SSO service builds a SAML assertion
    \item The browser issues an HTTP POST request to send the SAML to the SP
    \item Final access is allowed or denied to the user by the SP
\end{enumerate}

The opposite is a SP initiated SSO
\begin{enumerate}
    \item A user attempts to access a resource on the SP
    \item The SP sends a redirect in response 
    \item The SSO service checks for a pre-existing logon security context, if missing requires the user to
    provide credentials
    \item The user provides a valid credentials and a local security context is created
    \item The IdP creates a SAML assertion 
    \item The browser issues an HTTP POST request to the SP
    \item The access is allowed or denied to the user
\end{enumerate}

\subsubsection{Authentication Context}
It indicates how a user athenticated at an IdP. It can be required by an SP to enstablish a level of assurance
(LOA)

\subsubsection{Metadata}
A SAML metadata document describes a SAML deployment which is used to enstablish trust and interoperability.
The minimum required is the entity ID, cryptographic keys and protocols endpoints (bindings and URLs). It's 
important to remember that the keys are both used for encryption and digital signature. In this case is 
required that each party knows each other in advance.

\subsection{SAML Security}
Only providing assertion to enstablish trust may lead to man-in-the-middle or replay attacks. SAML fights this 
vulnerabilities by relieng on a Public Key infrastructure. It also uses SSL/TLS to grant authentication and 
confidentiality, and the use of XML Signature improves the message integrity.\\ Other mechanism used are
expiration timers and unique identifiers for the messages

regarding privacy instead, SAML uses persisten pseudonyms and one-time identifiers between an identity and a
service provider and creates a specific authentication context in which the user is allowed to use only certain 
operations

\subsection{National Identity Infrastructures - Spid}
The Spid is based on SAML 2.0 and is managed by the AGency for Digital Italy (AgID).\\The trust is achieved
thorugh the intermediation of the agency, a third party guarantor, a process of accreditation of digital 
identity providers, attribute authorities and service providers. 

All the entities that have passed the accreditation process are listed in the federation registry. For each record
is stored the SAML identifier, name of the subject, type of entity, URL to the service provider's metadata and a 
list of qualified attributes.

\subsubsection{CIE}
Used to store personal information, the eletronic Id Card is provided with a NFC cip with encryption

\subsection{European Identity Infrastructure}
An european service used for identification between nations of the EU. Uses various eIDAS-Nodes to connect the 
different country's services. The communication between a connector and a service relies on SAML

Currently EIDAS is under discussion with a focus on digital wallets, or a user controlled app that permits to 
select what personal information to share with a service. This gives self-sovereign identity, decentralized 
identifiers and verifiable credentials.\\Currently the EU is developing a European Self-Sovereign Identity Framework,
or ESSIF, with the use of decentralized identifiers and european blockchain services.

\newpage
\section{Access Control}
The access control is composed of various mechanism and actors, but we will focus on the subject, or entity, the 
access control module and the policy decision point (or PDP). This elements (with the exception of the subject)
are contained inside an outer and an inner boundary. 

Some remarks regarding the access control isolation boundaries: 
\begin{itemize}
    \item The outer boundary prevents the by-passing of the guard who authorizes all resources request
    \item the inner boundar protects the integrity of the audit log
\end{itemize}

NB: Principle of least priviledge: every subject must be able to access only the information and resources 
necessary for its legittimate purpose

NB: Need to know principle: a process should have access only to the needed object, been able to make only the needed
operation and only during the needed time

NB: Domain of protection: a set of objects and access rights

\subsubsection*{AC definition}
An AC is a process which grants or denies access to resources and data of a system. Its purpose is to protect 
resources from unauthorized access by using, depending on the context, various AC policies, or set of rules

\subsubsection*{Structured approach}
A structured approach to the AC divides policies a the higher level, followed by model and enforcement. In this case 
policies represent how they are specified, model models the meaning and enforcement how policies are implemented.

This method is usefull for implementing the use of automated reasoning to perform security analysis and to 
separate concerns

NB: Confused deputy: during a privilege escalation attack, the attacker does not have access to a resource, so 
it uses another subject, which has access, as an intermediary for read or modifying the data

\subsection{AC models}
\subsubsection{AC matrix}
The ACM lists all subjects and resources, and for each cell lists the permitted actions 

\subsubsection{Discretionary AC}
DAC policies are determinated by the owner user, this way no central entity is needed. This method is typically used 
in OSs.\\
To simplify administration, large organization arrange users in groups, and determine the policies based on the group
and not the single user🇮🇹

DAC offers the pros of flexibility and a more intuitive implementation. On the other hand, the protective state of 
resources is subjective to the owner and vulnerable to user's mistakes. DAC are also vulnerable to trojans, or 
malicious software unknowingly executed by users with their user priviledge

\subsubsection{Mandatory AC}
This policies are managed by a central identity which decides who can do what and on which resources. sharing
is enforced by the use of security labels and somethimes legal requirements.

All this points leads to a multi-level security model. This model has the objective of preventing the release of 
sensitive information to untrusted users.\\With this system, information is labeled with a sensitivity label, which 
joins a linear ordered set of sensitivity values (example: unclassified $\to$ confidential $\to$ topsecret) and the 
category to which the information belongs.\\When a resource is created, is the resource creator that applies a
sensitivity label bases on some criteria.\\
In multy level security, also the user is assigned clearance (or categorized) based on the security level and 
a set of need-to-know categories. 

After having defined sensitivity labels and clearances, the AC compares them and if the user has access to the data
based on the No Read Up and No Write Down properties: the first one declares that for accessing data you need to 
be on a higher or equal clearance level than the object. The second one instead says that higher level 


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\section{Wrap-up}
\subsection{Security in perspective}
\begin{itemize}
    \item Attacks violate one or more of the CIA properties
    \item Identifying which properties are violated help in
    \subitem understanding the impact and
    \subitem identifying which security services should be used to deploy security mechanisms for risk mitigation
    \item Ensuring security is a difficult task because of several issues
    \subitem attribution
    \subitem scale of and borderless attacks
    \subitem growing attack surface
    \subitem lack of collaboration among victims
    \item Security and privacy overlaps but are different and sometimes conflicting
    \item Privacy preserving techniques are relevant also from a legal point of view
    \subitem The GDPR encapsulates a risk based approach to privacy
    \item Security is not purely technical, human factors play a crucial role
    \subitem Hence the importance of security awareness and security training to improve the security posture
\end{itemize}
\subsection{Passwords and authentication}
\begin{itemize}
    \item Authentication amounts to verifying the identity of a user, process, or device, often
    as a prerequisite to allowing access to resources in an information system
    \item Here we focused on user authentication
    \item Passwords are one of the most widespread and accepted method for user
    authentication despite their shortcomings
    \item Adequate security mechanisms should be put in place to protect stored passwords
    \subitem Hashing, salting, access control
    \item Authentication is only one phase of the identity management lifecycle
    \subitem Also other phases (e.g., identity proofing) may have weaknesses
    \item Given the importance and difficulty of identity management, it is frequently
    outsourced to trusted third parties with advantages and disadvantages also from
    the viewpoint of security (and privacy)
\end{itemize}
\subsection{SAML}
\begin{itemize}
    \item SAML allows service providers to outsource identity management and focus on
    their core business
    \subitem Reduced administration burden
    \subitem Improved interoperability, usability, security and privacy
    \item SAML profiles are useful use case scenarios
    \subitem Web SSO most widely adopted
    \subitem SSO has increasing importance and is gaining wider and wider adoption
    \item SAML is ideal starting point to build infrastructures for digital identity management
    \subitem Key enablers in an increasing digital world (SPID, eIDAS)
    \subitem First line of defense against attackers
\end{itemize}


\end{document}
